Gaia's constraints on the extremely massive ($\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-10 VFTS682_APJL.tex
Selma sunday - mostly section 2 and tables

\documentclass[apjl,twocolumn]{emulateapj}
\usepackage{graphicx}
\usepackage[varg]{txfonts}
%\usepackage[pagebackref,breaklinks,colorlinks,citecolor=blue]{hyperref}        
\usepackage[%draft, 
pagebackref,colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
\renewcommand*{\backref}[1]{[#1]}  % gives neat back references in the pdf
\usepackage{aas_macros}
%\usepackage{amsmath} % clashes with emulateapjj for some reaons
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
%\usepackage{chngpage}
%\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
%\usepackage{forest}
%\usepackage{tikz-qtree}
%\usetikzlibrary{shadows,trees}

@@

\renewcommand{\labelitemii}{$\bullet$}
\newcommand{\kms}{{\,\mathrm{km\ s^{-1}}}}
\newcommand{\Msun}{{\,\mathrm{M}_\odot}}
\newcommand{\Lsun}{{\,\mathrm{L}_\odot}}
\newcommand{\masyr}{{\,\mathrm{mas}\,\mathrm{yr}^{-1}}}

@@

% \title{\emph{Gaia} DR2 identifies VFTS 682 as the most massive % , dynamically
%                                 % ejected %% to fit in one line
% runaway star known to date}
\title{Gaia's constraints on the  $\sim$$150\,M_\odot$  runaway star candidate VFTS682}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{4}$, J. Bestenlehner$^{5}$,
  V.~H\'enault-Brunet$^{6}$,  S.~Justham$^{1,7,8}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{9}$, H.~Sana$^{10}$, F.~R.~Schneider$^{11}$, J.~S.~Vink$^{12}$}
\affil{$^{1}${Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands} \\
  $^{2}$ {ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain} \\
  $^{3}$ {Space Telescope Science Institute, 3700 San Martin Drive,
    Baltimore, MD 21218, USA}\\
  $^{4}$ {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
  $^{5}${Department of Physics and Astronomy, Hicks Building,
    Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
  $^{6}$ {National Research Council, Herzberg Astronomy \&
    Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7,
    Canada}\\
  $^{7}$ {School of Astronomy \& Space Science, University of the Chinese
    Academy of Sciences, Beijing 100012, China}\\
  $^{8}$ {National Astronomical Observatories, Chinese Academy of
    Sciences, Beijing 100012, China}\\
  $^{9}$ {Argelander-Instit\"ut f\"ur Astronomie, Universit\"at Bonn,
    Auf dem H\"ugel 71, 53121, Bonn, Germany}\\
  $^{10}$ {Institute of Astronomy, KU Leuven, Celestijnenlaan 200 D, B-3001 Leuven, Belgium}\\
  $^{11}$ {Department of Physics, University of Oxford, Keble Road,
    Oxford OX1 3RH, UK} \\
  $^{12}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}\\
}

 \thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}

 How very massive stars form is still among the most intriguing open
 questions in stellar astrophysics.  VFTS 682 is among the most
 extreme massive stars known, with an inferred initial mass of about
 $150\,M_\odot$. It is located in 30 Doradus at a projected distance of 29\,pc from the central cluster R136.  The absence of other massive stars in its immediate vicinity lead to two intriguing possible hypotheses posed in earlier work. Either it formed in relative isolation through a new mode of star formation or it was ejected dynamically from the central cluster.

Aiming to shine light on this debate, we investigate the kinematics of  VFTS 682 as obtained by \emph{Gaia} and a \emph{Hubble Space Telescope} proper motion campaign.    We derive a projected velocity relative of to the cluster of $\sim$$30\pm20\kms$.  The direction and magnitude of the proper motion are consistent with ejection from the central cluster considering its inferred age.  However, the error bars and spread in proper motions derived for other sources in the region are substantial, which makes us cautious to derive hard conclusions.

If future data confirms the runaway nature, this would make the VFTS682 the most massive runaway star known to date. 
While we cannot prove this solidly from the current data, we do
consider this hypothesis plausible. The central cluster is known to
harbor several other stars more massive than $150\,M_\odot$ similar in
spectral type.  Moreover, the current record holders are O type stars
including VFTS 16 which provides direct evidence that R136 is indeed
capable of ejecting some of its most massive members. 
\end{abstract}

@@

One of the most extreme examples that has been considered in this debate is the very massive star VFTS682  \citep[][]{bestenlehner:11, bressert:12}. This star is located in the field of the 30 Doradus region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $\sim$$140^{+30}_{-16}\,M_\odot$ corresponding to an initial mass of $\sim$$150^{+30}_{-17}\,M_\odot$
\citep{schneider:18}. 
%
%$137.8^{+27.5}_{-15.9}\,M_\odot$,
%
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
%
From the spectral point of view, it is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
 in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
 %
 In particular, a remarkable similarity exist between the
spectrum of VFTS682 and R136a3 for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. R136 hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher.

@@

\begin{table}
  \begin{center}
    \caption{Observed and derived parameters for VFTS682. We also list
    the average proper motion components of R136 derived by
    \cite{lennon:18} for reference.}
    \begin{tabular}{ll|c|c}
      \hline
      \hline
      &Parameter & Value & Source\\

       \hline
       \multicolumn{4}{l}{\emph{Stellar Parameters VFTS 682}} \\
     & present day $M$ \hfill $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ &
                                                     \multirow{3}{*}{\cite{schneider:18}}\\
      &$M_\mathrm{ZAMS}$\hfill $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ &
      \\
      &age \hfill [Myr] & $1.0\pm0.2$ & \\[5pt]
       \multicolumn{4}{l}{\emph{Average Proper motion in the region}} \\         &$\langle\mu_\mathrm{RA}\rangle$\hfill[$\mathrm{mas\ yr^{-1}}$] & $1.74\pm0.01$
                        &  \multirow{2}{*}{\cite{lennon:18}}\\[5pt]
      &$\langle\mu_\mathrm{DEC}\rangle$\hfill[$\mathrm{mas\ yr^{-1}}$]
                & $0.70\pm0.02$ & \\
  %    \multicolumn{4}{l}{\emph{Position}} \\

    %  &RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
   %                     & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
    %  &DEC \hfill [degree] & -69.07 % $\pm$  0.05
     %                   & \\[5pt]
      \multicolumn{4}{l}{\emph{Relative Proper motion}} \\                 
      &$\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.10$\pm$0.07 & \multirow{2}{*}{\emph{Gaia} DR2}\\[5pt]
      &$\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.09 & \\[5pt]

      &$\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
      &$\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                        0.2$\pm$0.1 &
      \\[5pt]
      \multicolumn{4}{l}{\emph{\SdM{ Motion in polar coordinates }}} \\     
      &$\delta v_\mathrm{rad, R136}$  \hfill[$\kms$] & xxx & \cite{This Work}\\   
       &$\theta_\mathrm{rad, R136}$  \hfill[$\kms$] & xxx & \cite{This Work}\\

  %    &$\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\

@@

    \end{tabular}
    \tablecomments
    {The error on the RA and DEC positions, are of order
      $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$ in \emph{Gaia}
      DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

\subsection{ Overview of VFTS682 from previous studies \label{data:vfts683}}

The star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000 \citep[][% see also
% \Tabref{tab:vfts682} for the coordinates from \emph{Gaia} DR2
]{evans:11}
was originally classified as a young stellar object based on its
mid-infrared excess \citep{gruendl:09}. \citet{evans:11} reclassified the
object as Wolf-Rayet star of spectral type WNh5 using multi-epoch
spectra covering $\lambda$4000--7000 from the
VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.

Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.
\citet{bestenlehner:11} % further analyzed the spectra % using the non-LTE
% % model atmosphere code CMFGEN \citep{hillier:98}
also derived stellar
parameters and surface abundances. They inferred an extinction
of $A_V=4.45\pm0.12$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citet{schneider:18} estimated
a present-day mass for VFTS682 of $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$.% , using the Bayesian analysis tool BONNSAI
% \citep{schneider:17}, based on the evolutionary models from
% \citet{brott:11} and \cite{kohler:15}.
% These values place VFTS682 around the
% ``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

\citet{bestenlehner:11} reported for VFTS682 a wind mass loss rate of
$10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, % inferred from
% spectral analysis assuming a smooth outflow, i.e.,
not accounting for
clumping. The present-day surface helium mass
fraction is $Y=0.45\, (0.49)$ from the spectral analysis of
\cite{bestenlehner:11} (\citealt{rubio-diez:17}). This value can be
used to constrain the age of the star and its past evolution.

% The star is very similar from the spectral point of view to R136a3, located inside the star cluster
% R136 \citep{crowther:10}  for which \citet{crowther:16} report a
% current mass estimate of $180^{+30}_{-30}\Msun$. The R136 cluster hosts
% at least two more very massive WN5h stars, R136a1 and R136a2, whose
% estimated current masses are even higher. However, VFTS682 stands
% out by its isolation at a projected offset of 29\,pc from the center
% of the cluster \citep{bestenlehner:11}.

Further worth noticing is the variability of the
star. \citet{parker:93} reported B and V magnitudes roughly 0.5\,magnitude
lower than found by \citet{evans:11}.
\citet{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band at $\sim$10\% level on a timescale of years.
They commented that this is unusual for Wolf-Rayet stars and more reminiscent
of Luminous Blue Variable (LBV) stars. % \citep[e.g.,][]{langer:94}.

\cite{bressert:12} note that VFTS682 shows a significant radial
velocity offset compared to the nebular lines from the gas surrounding
it, suggesting peculiar motion in the line of sight direction.
\citet{bestenlehner:11} quote a radial velocity measured from the
N{\footnotesize V} $\lambda4944$ of   $300\pm10\kms$. The average radial velocity of the 30 Doradus region is 
$270\pm10\kms$. Whether this measurement truly reflects a peculiar radial velocity is debated. The determination of radial velocities for emission-lines stars is difficult because of the presence of an optically thick, non-homogeneous, and variable wind obscuring the photosphere. The radial velocity errors quoted above are only statistical, and do not quantify the (possibly larger)
systematic errors.  \todo{Possibly mention and cite paper in prep by Paco and student on Xshooter data?  Possibly mention (or even use) on their measurement for the RV?}

For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error on
the distance determination ($\lesssim2\%$) when converting proper motions into
physical velocities.
%We implicitly assume that the use of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not introduce significant errors.

@@

For large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even exceed \emph{Gaia} DR2 data in quality.

Recently, \citet{platais:18} presented HST astrometry and proper
motions in the 30 Doradus region (GO-12499; P.I.: D.~J.~Lennon). The
brightest stars (V$<$14) are missing from the catalogue because of
saturation, but the high extinction towards VFTS682 makes is faint
enough to be included. % in this catalogue.
The  proper motions are
listed in \Tabref{tab:vfts682}.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS682 relative to the field, providing a
measure of the motion of this star completely independent from
\emph{Gaia}. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$, which added in quadrature give a relative proper motion of
$\delta \mu^{HST} =0.20 \pm 0.15\,\mathrm{mas\ yr^{-1}}$.
% \begin{equation}
%   \label{eq:pm_around_HST}
%   \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
%   = 0.20 \pm 0.15\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}

>>>>

\documentclass[apjl,twocolumn]{emulateapj}
\usepackage{graphicx}
\usepackage[varg]{txfonts}
%\usepackage[pagebackref,breaklinks,colorlinks,citecolor=blue]{hyperref}        
\usepackage[%draft, pagebackref,
colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
%\renewcommand*{\backref}[1]{[#1]}  % gives neat back references in the pdf
\usepackage{aas_macros}
%\usepackage{amsmath} % clashes with emulateapjj for some reaons
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
%\usepackage{chngpage}
%\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
%\usepackage{forest}
%\usepackage{tikz-qtree}
%\usetikzlibrary{shadows,trees}

@@

\renewcommand{\labelitemii}{$\bullet$}
\newcommand{\kms}{{\,\mathrm{km\ s^{-1}}}}
\newcommand{\Msun}{{\,\mathrm{M}_\odot}}
\newcommand{\Lsun}{{\,\mathrm{L}_\odot}}
\newcommand{\Myr}{{\,\mathrm{Myr}}}
\newcommand{\masyr}{\,\mathrm{mas}\,\mathrm{yr}^{-1}}

@@

% \title{\emph{Gaia} DR2 identifies VFTS 682 as the most massive % , dynamically
%                                 % ejected %% to fit in one line
% runaway star known to date}
%\title{Gaia's constraints on the  $\sim$$150\,M_\odot$  runaway star candidate VFTS682}
\title{Gaia's constraints on the extremely massive  ($\sim$$150\,M_\odot$) runaway star candidate VFTS682}

\author{M.~Renzo$^{*,1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{4}$, 
  J.~Bestenlehner$^{5}$, \SdM{C.~J.~Evans$^{X}$},
  V.~H\'enault-Brunet$^{6}$,  S.~Justham$^{1,7,8}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{9}$,  \SdM{F. Najarro$^{X}$}, H.~Sana$^{10}$, F.~R.~Schneider$^{11}$, J.~S.~Vink$^{12}$}
\affil{ $^{*}$ Corresponding author.  A list of the affiliations can be found at the end of this paper.}

 \thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\begin{abstract}

 How very massive stars form is still among the most intriguing open  questions in stellar astrophysics.  VFTS 682 is among the most  extreme massive stars known, with an inferred initial mass of about  $150\,M_\odot$. It is located in 30 Doradus at a projected distance of 29\,pc from the central cluster R136.  The absence of other massive stars in its immediate vicinity led to two intriguing hypotheses posed in earlier work: either (a) it formed in relative isolation through a new mode of star formation or (b) it was ejected dynamically from the central cluster.

Aiming to shine light on this debate, we investigate the kinematics of  VFTS 682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space Telescope} photometry.    We derive a projected velocity relative of to the cluster of $\sim$$30\pm20\kms$.  The direction and magnitude of the proper is consistent with what is expected if VFTS 682 was ejected as a slow runaway from the central cluster, given its distance and the inferred age.  However, the error bars are substantial and the proper motion measurements alone cannot rule out the counter hypothesis with high confidence.

If future data confirms the runaway nature, this would make the VFTS682 the most massive runaway star known to date.  While we cannot prove this solidly from the current data alone, we do consider this hypothesis as the most plausible because of a variety of circumstantial clues. The central cluster is known to harbor several other stars more massive than $150\,M_\odot$ similar in spectral type.  The very massive O star VFTS 16, which has recently been identified as a dynamically ejected runaway, provides direct evidence that R136 is indeed capable of ejecting some of its most massive members, consistent with the predictions by numerical simulations.
\end{abstract}

@@

One of the most extreme examples that has been considered in this debate is the very massive star VFTS682  \citep[][]{bestenlehner:11, bressert:12}. This star is located in the field of the 30 Doradus region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $\sim$$140^{+30}_{-16}\,M_\odot$ corresponding to an initial mass of $\sim$$150^{+30}_{-17}\,M_\odot$
\citep{schneider:18}. 
%
%$137.8^{+27.5}_{-15.9}\,M_\odot$,
%
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
%
From the spectral point of view, it is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
 in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
 %
 In particular, a remarkable similarity exist between the
spectrum of VFTS682 and R136a3 \citep{rubio-diez:17} for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. R136 hosts
at least two more very massive WN5h stars, R136a1 and R136a2, whose
estimated current masses are even higher.

@@

\begin{table}
  \begin{center}
    \caption{Selected stellar parameters for VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}} \\
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ &
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ &
      \\
      &age & [Myr] & $1.0\pm0.2$ & \\
      &mass loss rate &  & $\SdM{xxx}$ & \\
      &helium abundance &  & $\SdM{xxx}$ & \\
      \hline

    \end{tabular}
    \tablecomments
    { 
    }
  \end{center}
  \label{tab:star}
\end{table}

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\

       \hline
           \multicolumn{5}{l}{\emph{Gaia's proper motion for VFTS 682 and average of the region}} \\
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & Correlation  ($\mu_\mathrm{RA}$, $\mu_\mathrm{DEC}$) &[$\masyr$] & $0.023$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (2) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (2)\\
\hline

  %    \multicolumn{5}{l}{\emph{Position}} \\

    %  &RA &[degree] &  \phantom{-}84.73 % $\pm$  0.03
   %                     & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
    %  &DEC & [degree] & -69.07 % $\pm$  0.05
     %                   & \\[5pt]
      \multicolumn{5}{l}{\emph{Gaia's relative proper motion for VFTS 682 }} \\                 
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & 0.10$\pm$0.07 & (1,4) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & 0.08$\pm$0.09 &  (1,4) \\
      &$\delta\mu_\mathrm{}$  &[$\masyr$] & \SdM{xxx}$\pm$\SdM{xxx} &  (1,4) \\
                 &$v_\mathrm{2d}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\  
 \hline     
          \multicolumn{5}{l}{\emph{HST's independent determination of the relative proper motion for VFTS 682}} \\  
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & 0.01$\pm$0.13 & (3) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & 0.2$\pm$0.1 &  (3) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & 0.20$\pm$0.15 &  (3) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & \SdM{$47\pm24$}\footnote{\SdM{This can't be right, the error bar is suddenly smaller than in the mas units!}} & (3)\\  
\hline
      \multicolumn{5}{l}{\emph{ Direction and magnitude w.r.t the vector pointing away from R136}} \\     
       &$\theta^{*}$  &[degrees] &  \SdM{xxx}$\pm$\SdM{xxx}  & (1,4)\\  
        &$v_\mathrm{2d, \parallel}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\  
 %          &$v_\mathrm{2d}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\

\hline
  %    &$\delta v_\mathrm{rad}$  &[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\

@@

    \end{tabular}
    \tablecomments
    { Proper motions are based on Gaia data unless indicated otherwise in subscript. The angle $\theta$ is defined such that $\theta=0$ corresponds to the direction pointing away from R136.
    $^1$\cite{Gaia-Collaboration+2018}, 
    $^2$\cite{lennon:18}, 
    $^3$\cite{platais:18},  
    $^4${\color{blue}This work}
    %The error on the RA and DEC positions, are of order $\sim$$10^{-2}\masyr$ in \emph{Gaia} DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

%\subsection{ Overview of VFTS682 from previous studies \label{data:vfts683}}

The star WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of the brightest stars in the region. The star is unlikely to have a close companion, because of the absence of clear radial velocity variations \citet{bestenlehner:11}.  Bayesian fits of the stellar parameters against evolutionary tracks \citep{brott:11, kohler:15} using the Bonsaii code \citep{schneider:17, schneider:18} provide estimates for the age, present mass and initial mass, indicating that this star among the most massive stars known, see Tab~ \label{tab:vfts682} for an overview of the parameters.

Estimates of the radial velocity are complicated by the variable, optically thick winds that is typical for emission line stars. \citet{bestenlehner:11} estimate a mass loss rate of $10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, not accounting for the possible effect of clumping.  The V-band light curve of VFTS682  shows variations at a $\sim$10\% level on a timescale of years, which is unusual for Wolf-Rayet stars and more typical for Luminous Blue Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also shows mid-infrared excess \citep{gruendl:09}.  We therefore caution for over-interpreting the estimates for the radial velocities, but we will summarize the estimates that have been published.  \citet{bestenlehner:11}  estimate a radial velocity (RV) of  $300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which is offset from the average radial velocity of the region, which is  $270\pm10\kms$. This is consistent with, but no proof of, a runaway nature.  \cite{bressert:12} note an offset compared between the RV of the star and the nebular lines from the gas filaments in its vicinity. This is consistent with, but also no proof of, the expectation if the star was not formed in situ.  Newer measurements based on new Xshooter data give different estimates for the RV ({\color{blue} Rubio-D{\' i}ez et al. in prep}). We will therefore refrain from using the RV measurements in this work.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.

%% Radial Velocity:

The distance to 30 Doradus is $50$\,kpc  \citep[][]{pietrzynski:13}. The error on the distance determination is small ($\lesssim2\%$) and the possible difference between VFTS682 and the average of the region is probably even smaller (likely $\ll 1\%$). These uncertainties are negligible compared to the errors in the proper motion discussed below.

%\citet{parker:93} reported B and V magnitudes roughly 0.5\,magnitude lower than found by \citet{evans:11}.  - Citing OGLE is enough

%\SdM{\citet{bestenlehner:11} further report a wind mass loss rate of $10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, without correcting for the possible effect of clumping.  The helium surface  abundance is  $Y=0.45\, (0.49)$ \cite{bestenlehner:11} (\citealt{rubio-diez:17}).  IN TABLE ONLY TO SAFE SPACE}

\subsection{ \emph{Gaia} astrometry for VFTS682\label{data:gaia}}

%\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}

VFTS682 is identified with the source id 4657685637907503744 in the \emph{Gaia} DR2 catalog  as a 15.65 mag star in the G band \citep{Gaia-Collaboration+2016a, Gaia-Collaboration+2018}.   The number of visibility periods, i.e. groups of observations separated from other groups by a gap of at least 4 days, used in the astrometric solution is 17 for this star. The reported astrometric excess noise is zero. \citep[][]{lindengren:18}.  These values suggest that the \emph{Gaia} DR2  data for VFTS682 are reliable. Generally the Gaia errors are correlated, but the correlation coefficient is small for this star, see Tab.~\ref{tab:vfts682}. We can therefore treat the proper motion components as uncorrelated.

To determine the relative proper motion with respect to R136, we follow  \citet{lennon:18} to define the motion of the local frame of reference using the average proper motion of nearby stars with reliable astrometric data.  They select  bright ($G<17$) stars within 0.05 degrees of R136 with and exclude sources with proper motion errors greater than $0.01\masyr$ in both co-ordinates (see their  Sect.~2.1 for further discussion).

With these we compute the relative proper motion $\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.  We also derive the angle $\theta$ between the direction of motion and the vector connecting the center of R136 with the current position of VFTS 682,  i.e. such that $\theta = 0$ is the angle expected for a star ejected from R136.    We further provide the 2D velocity projected onto the radial vector pointing away from R136 $v_\mathrm{2d, \parallel}$  and the total projected 2d velocity $v_\mathrm{2d}$.   All kinematic quantities are provided in Tab.~\ref{tab:vfts682}.

@@

%For large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even exceed \emph{Gaia} DR2 data in quality. \SdM{Mathieu: do you have a ref for this?}

The 30 doradus region was target of an a two-epoch photometric campaign with \emph{HST} providing observations in the F775W filter in October 2011 and October 2014 (GO-12499; P.I.: D.~J.~Lennon, \citealt{Sabbi+2013}). Visually, the star is clearly isolated and does not appear to have any close neighbor. This by itself gives further confidence in ability of \emph{Gaia} to measure the proper motion reliably without being affected by crowding.

 \citet{platais:18} recently analyzed the HST data for candidate runaway stars. The brightest stars (V$<$14) are missing from the catalogue because of saturation effects, but the high extinction towards VFTS682 makes it faint enough to be included.
 Our star did not make the more conservative selection criteria for runaway stars adopted in that study. However, the measurements are useful here, since they provides a measurement of the proper motion of VFTS 682 that is completely independent from the \emph{Gaia} measurement.

We list their data in \Tabref{tab:vfts682}.  The errorbars are larger than we obtain from \emph{Gaia}, but the values are consistent. \SdM{Actually they are not consistent with values found in one of Danny's emails. What is going on here?}

%ID     /HST relative mu_x/HST relative mu_y/HST error_mu_x/HST error_my_y/  HST F775W mag/Gaia B-mag?/Gaia relative mu_x/  Gaia relative mu_y/Gaia error_pmRA/Gaia error_pmDec

%682  -0.02   0.19 0.10 0.09 15.175 15.656   0.133   0.166 0.070   0.080

%The catalog from \citet{platais:18} lists the proper motion components of VFTS682 relative to the field, providing a
%measure of the motion of this star completely independent from
%\emph{Gaia}. They found proper
%motion components relative to the field
%$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
%  yr^{-1}}$ and
%$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
%  yr^{-1}}$, which added in quadrature give a relative proper motion of
%$\delta \mu^{HST} =0.20 \pm 0.15\masyr$.
% \begin{equation}
%   \label{eq:pm_around_HST}
%   \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
%   = 0.20 \pm 0.15\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}
